Zohmg---a large scale data store for aggregated time-series-based data. [DRAFT]

### 2009-08-12 doc/report/msc-report.tex
report: updated preface.

@@

[graphical element]

@@

This chapter will present the implementation of Zohmg. For a usage manual, see
the appendix.

@@

In the first section the implementation of project creation, configuration, and
setup will be presented. The second section features the data import
implementation. It will present example data as well as an example user map
function to show this. The third section presents the implementation of data
interchange between Zohmg and other software such as Hadoop, Dumbo, and HBase.
It also shows an overall of Zohmg. The third section also presents the
implementation details of Zohmg's data export component.

@@

Zohmg employes the concept of projects. A project is a self contained
application data directory. A Zohmg project is necessary for using Zohmg; they
are created with a management script, which is also used for interacting with
the Zohmg system. When the user has created a Zohmg project directory further
actions can be performed. The Zohmg projects reside within a project directory,
which contains five directories: \texttt{clients}, \texttt{config},
\texttt{lib}, and \texttt{mappers}. The details of these directories are
presented throughout this chapter.

@@

Zohmg needs configuration of the user's data sets and the environment for Hadoop
and Dumbo. When the data set files are configured, the user runs the management
script, which executes the configuration and creates tables in HBase according
to the configuration.

@@

A data set is any set of data that the user has. Zohmg requires the user to
write a configuration that describes the attributes of the data. The
configuration of a data set is a list of dimensions and projections. This
configuration will be read during the import phase to match analysed data
against the schema definition.

The data set configuration files reside in the project's \texttt{config}
directory. The data sets are configured using YAML\footnote{YAML Ain't a Markup
Language}, a human-readable data serialization format. Any human-readable
configure file method could have been chosen. YAML was chosen because it is is
very simple for humans to read and edit, and likewise for machines. The Python
module PyYAML was used to implement the YAML configuration reading part.
\cite{pyyaml}

A typical configuration looks like this:

\begin{lstlisting}[caption=A data set configuration example.,captionpos=b]
dataset: submissions

dimensions:
 - artist
 - track
 - user

projections:
 - artist
 - user
 - user-artist
 - user-track
 - artist-track

units:
 - scrobbles
 - loves
 - bans
 \end{lstlisting}

@@

The execution environment is described by a Python script. In it are defined
run-time variables which are required by Hadoop and Dumbo, namely the Java class
path (CLASSPATH) and the absolute path to the Hadoop directory (HADOOP_HOME).

By using a Python script for run-time variables leaves the problem of validation
to Python. The environment configuration file resides in the project's
\texttt{config} directory.

@@

In order to import data to Zohmg, the user writes a map function in Python.
Zohmg employs Dumbo to run the map function on the Hadoop instance defined by
the HADOOP_HOME variable.

A standard MapReduce map function emits key-value pairs. The Zohmg user map
function emits triples (tuples with three values), which consist of a
time-stamp, a hash table of dimensions and their respective values, and a hash
table with units and their respective measurements (values). In effect, the
dimensions and the time-stamp specify a point in n-space (the key) and the
measurements represent the observations made at that point (the value(s)).

A reducer function that is specific to Zohmg is employed to perform aggregation
on the output from the map phase. The reducer sums the measurements for each
point in n-space and for each unit and emits the resulting aggregates.

The output from the reducer is interpreted by a custom output reader. Commonly,
an output reader stores the output of the reduce phase on disk. The output
reader that Zohmg employs instead stores the aggregates in HBase. It is written
in Java.

\subsection*{Example of Importing}

@@

\begin{lstlisting}[language=Python,caption={An example map function, parsing
Apache access logs.},captionpos=b]
   def map(key, value):
      log = parse_apache_log(value)
      dimensions = {'useragent' : log.useragent,
                    'path'      : log.path,
                    'status'    : log.statuscode}
      measurements = {'pageviews' : 1,
                      'bytes'     : log.size}

      yield log.timestamp, dimensions, measurements
\end{lstlisting}

@@

\begin{verbatim}
   zohmg import mappers/apache.py /data/weblogs/2009/04/20
\end{verbatim}

@@

The underlying data store is HBase, which is a column-oriented database. Column
families are pre-defined, but everything else, such as column qualifiers and
rowkeys, can assume arbitrary values.

Zohmg models the aggregated data with this in mind. An HBase table is created
for each data set, and in the table is a single column family called 'unit'. For
every unit that the data set contains there will be a column qualifier. For
example, 'unit:pageviews' would be one such column qualifier in the example
above. Employing the qualifier to encode arbitrary unit names means that the
user can add new units to his data set configuration without having to change
the schema of the HBase table.

Zohmg stores one or more projections of the data, and each projection is made up
of one or more dimensions in addition to the ever-present time dimension. Every
dimension is a tuple of dimension name and dimension value. The dimension tuples
of the projection are stored in the rowkey. The rowkey is a string; the
dimension names and values are separated by a dash. As such, the dash is one of
the few disallowed characters in dimension names and values.

For example, a rowkey for the projection of the user agent and status code
dimensions might look like this:

\begin{lstlisting}[caption=Example of rowkeys,captionpos=b]
    useragent-firefox-statuscode-200-20090420
    useragent-firefox-statuscode-200-20090421
    useragent-firefox-statuscode-200-20090422
    ..
    useragent-firefox-statuscode-404-20090420
    useragent-firefox-statuscode-404-20090421
    useragent-firefox-statuscode-404-20090422
    ..
    useragent-safari-statuscode-200-20090420
    useragent-safari-statuscode-200-20090421
    ..
\end{lstlisting}

The rowkeys are sorted lexicographically. This means that all entries for the
user agent 'firefox' will be found close to each other. If the data analyst is
more interested in finding all entries with status code 200, it would make sense
to set up the projection so that status code is the first dimension.

@@

The reducer needs to communicate to the OutputReader under what row key,
column-family and qualifier to store the computed sum. It does this by
emitting the row key as the key and a JSON-serialized map of the form \\
\texttt{\{"column-family:qualifier" : value\}}.

@@

% TODO: not truly correct
% TODO: A detailed explanation of generally everything in the darling dir.
The Darling output format for HBase. Between Dumbo, which Zohmg uses for data
import, and JSON is used for serializing data.

>>>>

\vspace{48pt}

\noindent {\Large \textbf{Sammanfattning}}
\vspace{12pt}

% TODO: sync with english abstract.

\noindent Att analysera data p{\aa} en massiv skala {\"a}r en av de st{\"o}rsta
utmaningarna som Last.fm st{\aa}r inf{\"o}r. Att tolka m{\"o}nster i
anv{\"a}ndarbeteende blir en utmaning n{\"a}r miljoner anv{\"a}ndare samspelar i
miljarder kombinationer. Datam{\"a}ngderna m{\aa}ste analyseras, summeras och
presenteras visuellt.

Detta examensarbete beskriver ett datalager f{\"o}r multidimensionell
tidsseriebaserad data. M{\aa}tt {\"a}r summerade {\"o}ver multipla dimensioner.
Datalagret {\"a}r optimerat f{\"o}r dataextraheringshastighet: Ett av
designm{\aa}len {\"a}r att servera data i musklickshastighet f{\"o}r att
fr{\"a}mja utforskning av data i realtid.

Liknande datalager existerar men de anv{\"a}nder oftast relationella
databassystem som databas f{\"o}r back-end. Nyheten i v{\aa}rt angripss{\"a}tt
{\"a}r att modellera multidimensionella datakuber ovanp{\aa} en distribuerad,
kolumnorienterad databas f{\"o}r att utnyttja skalbarhetsf{\"o}rdelarna av
s{\aa}dana databaser.

\pagebreak
\phantom{ }
\pagebreak

\noindent {\Large \textbf{Preface}}
\vspace{12pt}

\noindent Per Andersson and Fredrik M{\"o}llerstrand both study Computer
Science at Chalmers University of Technology in Gothenburg, Sweden. This thesis
is part of their master's degree. The work was performed at Last.fm's
office in London during spring 2009.

The authors would like to especially thank the supervisors---David Sands at
Chalmers and Johan Oskarsson and Martin Dittus at Last.fm---for their help and
support.

Per would like to especially thank Hedvig Kamp for help and guidance with the
thesis and general matters in life.

The authors would also like to thank their respective families and friends, and
the staff at Last.fm for their moral support before, during, and after the
thesis work.

\vspace{24pt}

\noindent \textbf{Acknowledgements}
\vspace{12pt}

\noindent Figure 2.1 is Copyright {\copyright} 2008 Robert Burrell Donkin,
licensed under Creative Commons Attribution 3.0 License.

\pagebreak
\phantom{ }
\pagebreak

@@

\begin{figure}[h]
    \begin{center}
        \includegraphics[scale=0.50]{map-reduce.eps}
        \caption{Overview of MapReduce dataflow.}
    \end{center}
\end{figure}

@@

This chapter will present the implementation of the Zohmg system. For a usage
manual see the appendix.

@@

In the first section the implementation of project creation, configuration, and
setup, will be presented. The second section features the data import
implementation; it will present example data as well as an example user map
function to import this. The third section presents the implementation of data
interchange between Zohmg and other software, such as Hadoop, Dumbo, and HBase,
but also throughout the Zohmg system. The third section also presents the
implementation details of the Zohmg system's exporting component.

@@

Zohmg uses the concept of projects. These projects are self contained
application data directories which are used with Zohmg. A Zohmg project is
necessary for using Zohmg; they are created with a management script, which is
also used for interacting with the Zohmg system. When the user has created a
Zohmg project directory further actions can be performed. The Zohmg projects
reside within a project directory, which contains five directories:
\texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers}, and
\texttt{transformers}. The details of these directories are presented throughout
this chapter.

@@

The Zohmg system needs configuration of the user's data sets and the environment
for Hadoop and Dumbo. When the data set files are configured, the user runs the
management script, which executes the configuration and creates tables in HBase
according to the configuration.

@@

% TODO: define what a data set is, and what makes it different from a project.

In this context a data set is the schema that the user defines for wanted data
sets to be handled by Zohmg. The configuration of a data set is simply a list
of dimensions and projections, which will layout the HBase tables according to
the schema. This configuration will also be read during the import phase, for
handling data to be imported.

The data set files reside in the project's \texttt{config} directory. The data
sets are configured using YAML\footnote{YAML Ain't a Markup Language}, a
human-readable data serialization format. Any human-readable configure file
method could have been chosen. YAML was chosen because it is is very simple
for humans to read and edit, and likewise for machines. The Python module
PyYAML was used to implement the YAML configuration reading part. \cite{pyyaml}

@@

The environment is a Python script which defines run-time variables, required by
Hadoop and Dumbo. Using a Python script for run-time variables leaves the
problem of validation to Python. The environment configuration file resides in
the project's \texttt{config} directory.

@@

In order to import data to Zohmg, the user writes a map function in Python which
Zohmg wraps; interpreting the output from the mapper program and storing it in
HBase.

A standard MapReduce map function emits key-value pairs. The Zohmg user map
function, instead, emits triples (tuples with three values), which consist of a
time-stamp, a hash table of dimensions and their respective values, and a hash
table with measurements (values). In effect, the dimensions and the time-stamp
specify a point in n-space and the measurements represent the observations made
at that point.

@@

\begin{lstlisting}[language=Python,caption={An example map function, parsing
Apache access logs.},captionpos=b]
   def map(key, value):
      log = parse_apache_log(value)
      dimensions = {'useragent' : log.useragent,
                    'path'      : log.path,
                    'status'    : log.statuscode}
      values = {'pageviews' : 1,
                'bytes'     : log.size}

      yield log.timestamp,dimensions,measurements
\end{lstlisting}

@@

\begin{verbatim}
   zohmg import mappers/apache.py \
   /data/weblogs/2009/04/20
\end{verbatim}

@@

% TODO: renskriv.
The typical use case for Zohmg is to read one or more column-qualifiers,
which correspond to one or more points in n-space, over a range of time.
Therefor, the keys are made up of the time and unit of the measurement,
and the qualifier queried for represents the point in n-space. More on
this in the data model section.

Explain our data model, the relation between column-families, column-qualifiers
and projections and dimensions.

@@

% TODO: data interchange.
The reducer needs to communicate to the OutputReader under what row key,
column-family and qualifier to store the computed sum. It does this by
emitting the row key as the key and a JSON-serialized map of the form \\
\texttt{\{"column-family:qualifier" : value\}}.

@@

% TODO: not truly correct
% TODO: A detailed explanation of generally everything in the darling dir.
The Darling output format for HBase. Between Dumbo, which Zohmg uses for data
import, and Hadoop JSON is used for serializing data.
